---
title: "Computing assignment 2"
author: "Alyssa Vanderbeek"
date: "21 October 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
```

## Problem 1

```{r 3plus3_function}
# p.true = vector of true DLT probabilities for each test dose
# n.dose = number (integer) of test doses
# N = number of simulations to run
performance_3plus3 <- function(p.true, n.dose, N){
  results = vector(mode = "list", length = N)
  
  for (sim in 1:N) {
    pt = 1 # patient on dose k
    n = 1 # patient on trial
    dose = 1 # dose k
    esc = 1 # escalation yes/no
    mtd = 0 # MTD
    
    dlts = vector(mode = 'list', length = n.dose)
    
    while (esc == 1 && dose <= n.dose && pt <= 6) {
      dlts[[dose]][pt] = rbinom(1, 1, p.true[dose])
      
      # If <3 pts enrolled on current dose and <2 DLTs, enroll next pt on same dose
      if (sum(dlts[[dose]]) <= 1 && length(dlts[[dose]]) < 3) {
        pt = pt + 1
        n = n + 1
        dose = dose
        # If 1/3 pts on current dose observe DLT, enroll next pt on same dose
      } else if (sum(dlts[[dose]]) == 1 && length(dlts[[dose]]) == 3) {
        pt = pt + 1
        n = n + 1
        dose = dose
        # If 0/3 pts on current dose observe DLT, escalate dose
      } else if (sum(dlts[[dose]]) == 0 && length(dlts[[dose]]) == 3) {
        pt = 1
        dose = dose + 1
        n = n + 1
        # If 2 patients out of 2-3 observe DLT, stop trial and declare the previous dose MTD
      } else if (sum(dlts[[dose]]) == 2 && length(dlts[[dose]]) <= 3) {
        esc = 0
        mtd = dose - 1
        # If <=1 patient out of 4-6 observe DLT, stay at same dose
      } else if (sum(dlts[[dose]]) == 1 && length(dlts[[dose]]) > 3 && length(dlts[[dose]]) < 6) {
        pt = pt + 1
        n = n + 1
        dose = dose
        # If we observe 1/6 DLTs, stop escalation and declare this dose level the MTD
      } else if (sum(dlts[[dose]]) == 1 && length(dlts[[dose]]) == 6) {
        esc = 0 # MTD
        mtd = dose
        # If we observe 2 DLTs for 4-6 pts, stop the trial and declare the previous dose the MTD
      } else if (sum(dlts[[dose]]) == 2 && length(dlts[[dose]]) > 3 && length(dlts[[dose]]) <= 6) {
        esc = 0
        mtd = dose - 1
        # In case of error
      } else {
        dlts[[dose]][pt] = NA # reassign entry as NA in case of error
      }
    }
    
    results[[sim]] = list(dlts = unlist(lapply(dlts, sum)),
                          nperdose = unlist(lapply(dlts, length)),
                          n = n,
                          mtd = mtd)
  }
  return(results)
}
```


```{r p1}

pcs.33 = c()

p.true1 = c(0.017, 0.043, 0.10, 0.22, 0.41)
n.dose = 5
N = 1e3
p = 0.1

set.seed(1)
sims = performance_3plus3(p.true1, n.dose, N)

n.dlts = do.call(rbind, lapply(sims, `[[`, 1))
n.pts = do.call(rbind, lapply(sims, `[[`, 2))
mtd.1 = unlist(lapply(sims, `[[`, 4))

mtd.selection = table(mtd.1[which(mtd.1 != 0)])/N
if (length(mtd.selection) < 5) {
  r = 5 - length(mtd.selection)
  mtd.selection = c(mtd.selection, rep(0, r))
}

pcs.33[1] = mtd.selection[which(p.true1 == p)]

## Table of performance metrics
rbind(p.true1,
      mtd.selection, #table(mtd.1)[-1]/N,
      colMeans(n.pts),
      colMeans(n.dlts)) %>%
  `rownames<-`(c("True DLT probability",
                             "Selection probability",
                             "Avg number of patients treated",
                             "Avg number of patients with DLT")) %>%
  knitr::kable(row.names = TRUE)
```


## Problem 2

```{r 2a}
p.true2 = c(0.10, 0.22, 0.41, 0.64, 0.81)
n.dose = 5
N = 1000

set.seed(1)
sims = performance_3plus3(p.true2, n.dose, N)

n.dlts = do.call(rbind, lapply(sims, `[[`, 1))
n.pts = do.call(rbind, lapply(sims, `[[`, 2))
mtd.1 = unlist(lapply(sims, `[[`, 4))

mtd.selection = table(mtd.1[which(mtd.1 != 0)])/N
if (length(mtd.selection) < 5) {
  r = 5 - length(mtd.selection)
  mtd.selection = c(mtd.selection, rep(0, r))
}

pcs.33[2] = mtd.selection[which(p.true2 == p)]

## Table of performance metrics
rbind(p.true2,
      mtd.selection,
      colMeans(n.pts),
      colMeans(n.dlts)) %>%
  `rownames<-`(c("True DLT probability",
                 "Selection probability",
                 "Avg number of patients treated",
                 "Avg number of patients with DLT")) %>%
  `colnames<-`(1:n.dose) %>%
  knitr::kable(row.names = TRUE,
               caption = "Performance metrics for 3+3 design when true DLT rates are p=0.10, 0.22, 0.41, 0.64, 0.81")
```

```{r 2b}
p.true3 = c(0.043, 0.10, 0.22, 0.41, 0.64)
n.dose = 5
N = 1000

set.seed(1)
sims = performance_3plus3(p.true3, n.dose, N)

n.dlts = do.call(rbind, lapply(sims, `[[`, 1))
n.pts = do.call(rbind, lapply(sims, `[[`, 2))
mtd.1 = unlist(lapply(sims, `[[`, 4))

mtd.selection = table(mtd.1[which(mtd.1 != 0)])/N
if (length(mtd.selection) < 5) {
  r = 5 - length(mtd.selection)
  mtd.selection = c(mtd.selection, rep(0, r))
}

pcs.33[3] = mtd.selection[which(p.true3 == p)]

## Table of performance metrics
rbind(p.true3,
      mtd.selection,
      colMeans(n.pts),
      colMeans(n.dlts)) %>%
  `rownames<-`(c("True DLT probability",
                 "Selection probability",
                 "Avg number of patients treated",
                 "Avg number of patients with DLT")) %>%
  `colnames<-`(1:n.dose) %>%
  knitr::kable(row.names = TRUE,
               caption = "Performance metrics for 3+3 design when true DLT rates are p=0.043, 0.10, 0.22, 0.41, 0.64")
```


```{r 2c}
p.true4 = c(0.007, 0.017, 0.043, 0.10, 0.22)
n.dose = 5
N = 1000

set.seed(1)
sims = performance_3plus3(p.true4, n.dose, N)

n.dlts = do.call(rbind, lapply(sims, `[[`, 1))
n.pts = do.call(rbind, lapply(sims, `[[`, 2))
mtd.1 = unlist(lapply(sims, `[[`, 4))

mtd.selection = table(mtd.1[which(mtd.1 != 0)])/N
if (length(mtd.selection) < 5) {
  r = 5 - length(mtd.selection)
  mtd.selection = c(mtd.selection, rep(0, r))
}

pcs.33[4] = mtd.selection[which(p.true4 == p)]

## Table of performance metrics
rbind(p.true4,
      mtd.selection,
      colMeans(n.pts),
      colMeans(n.dlts)) %>%
  `rownames<-`(c("True DLT probability",
                 "Selection probability",
                 "Avg number of patients treated",
                 "Avg number of patients with DLT")) %>%
  `colnames<-`(1:n.dose) %>%
  knitr::kable(row.names = TRUE,
               caption = "Performance metrics for 3+3 design when true DLT rates are p=0.007, 0.017, 0.043, 0.10, 0.22")
```



```{r 2d}
p.true5 = c(0.003, 0.007, 0.017, 0.043, 0.10)
n.dose = 5
N = 1000

set.seed(1)
sims = performance_3plus3(p.true5, n.dose, N)

n.dlts = do.call(rbind, lapply(sims, `[[`, 1))
n.pts = do.call(rbind, lapply(sims, `[[`, 2))
mtd.1 = unlist(lapply(sims, `[[`, 4))

mtd.selection = table(mtd.1[which(mtd.1 != 0)])/N
if (length(mtd.selection) < 5) {
  r = 5 - length(mtd.selection)
  mtd.selection = c(mtd.selection, rep(0, r))
}

pcs.33[5] = mtd.selection[which(p.true5 == p)]

## Table of performance metrics
rbind(p.true5,
      mtd.selection,
      colMeans(n.pts),
      colMeans(n.dlts)) %>%
  `rownames<-`(c("True DLT probability",
                 "Selection probability",
                 "Avg number of patients treated",
                 "Avg number of patients with DLT")) %>%
  `colnames<-`(1:n.dose) %>%
  knitr::kable(row.names = TRUE,
               caption = "Performance metrics for 3+3 design when true DLT rates are p=0.003, 0.007, 0.017, 0.043, 0.10")

mean_n = mean(unlist(lapply(sims, `[[`, "n")))
```


## Problem 3

```{r crm, results='hide'}
library(dfcrm)

crm_wrapper <- function(prior, truth, n, target, N, guess) {
  tru = which(truth == target)
  p0 = prior
  sim = crmsim(truth, prior = p0, target = target, n = n, x0 = 3, nsim = N)
  
  return(c(truth = tru, sim))
}

ppp = list(p.true1, p.true2, p.true3, p.true4, p.true5)

priors = lapply(1:5, function(i){
  getprior(halfwidth = 0.05, target = 0.1, nu = i, nlevel = n.dose)
})

# across all truths, how does our skeleton perform?
test = lapply(priors, function(p){
  lapply(ppp, function(i){
    set.seed(p)
    crm_wrapper(prior = p, truth = i, n = 31, target = 0.1, N = 50, guess = i)
  })
})

truMTD = lapply(test, function(i){
  lapply(i, `[[`, "truth")
}) %>%
  unlist %>%
  unique

selection_prob = lapply(1:length(test), function(i){ # for each prior
  lapply(1:length(test[[i]]), function(j){
    test[[i]][[j]]$MTD[truMTD[j]] # get the probability of selecting the correct dose as MTD under different truths
  }) %>% unlist
})

PCS.crm = lapply(selection_prob, mean) %>% unlist

```

Compared to the 3+3, the CRM is more consistent (less variable) in its selection of MTD. It therefore enrolls more patients at or near the target toxicity level (TTL), whereas the 3+3 does not allow specification of a TTL. But this comes at the cost of a larger overall sample size; n=31 in the CRM compared to an average N of `r mean_n` in the 3+3. 

Under all 5 true scenarios, I establish the skeleton as `r round(priors[[which(PCS.crm == max(PCS.crm))]], 3)` (I guess that the true MTD is the 4th dose, with halfwidth of 0.05). The average probability of correct selection (PCS) under this prior is `r PCS.crm[[which(PCS.crm == max(PCS.crm))]]`.


```{r crm_metrics}

optimal = test[[4]]
## Table of performance metrics
rbind(p.true1,
      optimal[[1]]$MTD,
      optimal[[1]]$level,
      optimal[[1]]$tox) %>%
  `rownames<-`(c("True DLT probability",
                 "Selection probability",
                 "Avg number of patients treated",
                 "Avg number of patients with DLT")) %>%
  `colnames<-`(1:n.dose) %>%
  knitr::kable(row.names = TRUE,
               caption = "Performance metrics for 3+3 design when true DLT rates are p=0.017, 0.043, 0.10, 0.22, 0.41")

## Table of performance metrics
rbind(p.true2,
      optimal[[2]]$MTD,
      optimal[[2]]$level,
      optimal[[2]]$tox) %>%
  `rownames<-`(c("True DLT probability",
                 "Selection probability",
                 "Avg number of patients treated",
                 "Avg number of patients with DLT")) %>%
  `colnames<-`(1:n.dose) %>%
  knitr::kable(row.names = TRUE,
               caption = "Performance metrics for 3+3 design when true DLT rates are p=0.10, 0.22, 0.41, 0.64, 0.81")

## Table of performance metrics
rbind(p.true3,
      optimal[[3]]$MTD,
      optimal[[3]]$level,
      optimal[[3]]$tox) %>%
  `rownames<-`(c("True DLT probability",
                 "Selection probability",
                 "Avg number of patients treated",
                 "Avg number of patients with DLT")) %>%
  `colnames<-`(1:n.dose) %>%
  knitr::kable(row.names = TRUE,
               caption = "Performance metrics for 3+3 design when true DLT rates are p=0.043, 0.10, 0.22, 0.41, 0.64")

## Table of performance metrics
rbind(p.true4,
      optimal[[4]]$MTD,
      optimal[[4]]$level,
      optimal[[4]]$tox) %>%
  `rownames<-`(c("True DLT probability",
                 "Selection probability",
                 "Avg number of patients treated",
                 "Avg number of patients with DLT")) %>%
  `colnames<-`(1:n.dose) %>%
  knitr::kable(row.names = TRUE,
               caption = "Performance metrics for 3+3 design when true DLT rates are p=0.007, 0.017, 0.043, 0.10, 0.22")

## Table of performance metrics
rbind(p.true5,
      optimal[[5]]$MTD,
      optimal[[5]]$level,
      optimal[[5]]$tox) %>%
  `rownames<-`(c("True DLT probability",
                 "Selection probability",
                 "Avg number of patients treated",
                 "Avg number of patients with DLT")) %>%
  `colnames<-`(1:n.dose) %>%
  knitr::kable(row.names = TRUE,
               caption = "Performance metrics for 3+3 design when true DLT rates are p=0.003, 0.007, 0.017, 0.043, 0.10")

```