---
title: "Computing Assignment 1"
author: "Alyssa Vanderbeek"
date: "9/29/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

source("~/Desktop/FALL 2019/Bayesian Design/P8133_BayesianAnalysis_AdaptiveDesign/fns.R")
```

## Question 1


```{r 2stg_adaptive}
p0 = 0.25
p1 = 0.4

n1 = 20 # stage 1 sample size
s1 = 6 # stop for futility after stage 1 if less than s1 patients respond
ntotal = 71 # total trial sample size (n1 + n2)
s2 = 24 # "go" decision if at least s2/ntotal patients respond

# Probability of a "go" decision under null
# t = c()
# for (i in s1:n1) {
#   e1 = dbinom(i, n1, p0) # Pr(s1 = i | p = p0)
#   e2 = 0
#   if (s2 - i > 0) {
#     for (j in (s2 - i):(ntotal - n1)) {
#       e2 = e2 + dbinom(j, ntotal - n1, p0) # sum {Pr(s1 >= s2-i | p = p0)}
#     }
#   } else {
#     for (j in 0:(ntotal - n1)) {
#       e2 = e2 + dbinom(j, ntotal - n1, p0)
#     }
#   }
#   t[i] = e1*e2
# }
# pgo.2stg = sum(t, na.rm = TRUE)
# 
# pnogo.2stg = 1 - pgo.2stg # Pr(no go) = 1 - Pr(go) = 0.951

# pnogo.2stg = 1 - go.2stg(n1, s1, ntotal - n1, s2, p0)
```

```{r Bayesian_nogo}
### Prior distribution of null and alternative response rates
n = 71 # total sample size
s = 24 # futility interim
N = 10e4

a.s = 25 # prior a for control 
b.s = 75 # prior b for control

a = 0.5
b = 1.5
a_cond = a + s # prior a for exp conditional on s
b_cond = b + n - s # prior b for exp conditional on s

ps = rbeta(N, a.s, b.s) # simulate control responses from prior
pe = rbeta(N, a_cond, b_cond) # simulate exp responses from prior


### Grid search for delta and alpha
d = 0.15 # seq(0, 0.2, 0.01)[-1]
alpha = seq(0, 1, 0.05)[-1]


t = sum(pe > (ps + d)) / N # left side of Bayesian decision rule
# take only those values of alpha for which s=22 lends a "go" decision
alpha = alpha[(which((t > alpha) == TRUE))]; alpha 

## go/no-go decision for s=24
t1 = sum(pe > (ps + d)) / N 

## go/no-go decision for s=23
s = s - 1
a_cond = a + s
b_cond = b + n - s

pe = rbeta(N, a_cond, b_cond)
t2 = sum(pe > (ps + d)) / N; t2 > alpha 

## Verify that the selected alpha makes the decision boundary s = 24
index = intersect(which((t1 > alpha) == TRUE), which((t2 > alpha) == FALSE))
alpha.test = alpha[index] # new alphas

# This should be a "go" decision
s = s + 1
a_cond = a + s
b_cond = b + n - s
pe = rbeta(N, a_cond, b_cond)
t1 = sum(pe > (ps + d)) / N; t1 > alpha.test # TRUE -> go

# This should be a "no-go" decision
s = s - 1
a_cond = a + s
b_cond = b + n - s
pe = rbeta(N, a_cond, b_cond)
t2 = sum(pe > (ps + d)) / N; t2 > alpha.test # FALSE -> no-go

```


The two-stage adaptive design as described in lecture is as follows: stage 1 enrolls 20 patients. At the interim, we stop the trial and conclude futility if less than 6 subjects respond. Otherwise, we continue onto stage 2 and enroll an additional 51 subjects for a total sample size of 71 subjects. At the end of the trial, we conclude futility ("no-go") if fewer than 24 subjects respond, and conclude efficacy ("go") if at least 24 subjects respond. 

To mimic this design in the Bayesian formulation, we can set $(\delta, \alpha)$ = (`r d`, `r alpha.test`), which gives a "go" decision for 24 responses, and a no-go decision for 23 responses.


